Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-16 talk/iwtc11/paper.tex
killed inline

Loop peeling is achieved by appending an inlined copy of the traced iteration at
the end of itselfe. See Figure~\ref{fig:overview}.
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration.

@@

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the preamble, $J$, and jump
arguments $K$. Looking  at the peeled version of our example in Figure~\ref{fig:peeled-trace} we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

After that, the peeled loop
will no longer be in SSA form as it operates on $i_3$ and $i_4$
which are not part of it. The solution is to extend the input
arguments, $J$, with those two variables. This will also extend the
jump arguments of the preamble, which is also $J$. 
Implicitly that also extends the jump arguments of the peeled loop, $K$,
since they are the inlined versions of $J$. For the example $I$ has to
be replaced by $\hat I$ which is formed as a concatenation of $I$ and
$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
$\hat K$ which is formed as a concatenation of $K$ and 
$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$. XXX: Maybe we should
replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?

>>>>

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itselfe. See Figure~\ref{fig:overview}.
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintian the SSA-property.

@@

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the preamble, $J$, and jump
arguments $K$. Looking  at the peeled version of our example in Figure~\ref{fig:peeled-trace} we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second copy of the trace (the peeled loop) from the
first (the preeamble) we need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
copying. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

After that, the peeled loop
will no longer be in SSA form as it operates on $i_3$ and $i_4$
which are not part of it. The solution is to extend the input
arguments, $J$, with those two variables. This will also extend the
jump arguments of the preamble, which is also $J$. 
Implicitly that also extends the jump arguments of the peeled loop, $K$,
since they are the image of $J$ under $m$. For the example $I$ has to
be replaced by $\hat I$ which is formed as a concatenation of $I$ and
$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
$\hat K$ which is formed as a concatenation of $K$ and 
$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$. XXX: Maybe we should
replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?
